% Copyright 2011-2012 David Hadka.  All Rights Reserved.
%
% This file is part of the MOEA Framework User Manual.
%
% Permission is granted to copy, distribute and/or modify this document under
% the terms of the GNU Free Documentation License, Version 1.3 or any later
% version published by the Free Software Foundation; with the Invariant Section
% being the section entitled "Preface", no Front-Cover Texts, and no Back-Cover
% Texts.  A copy of the license is included in the section entitled "GNU Free
% Documentation License".

\chapter{Introduction}

The MOEA Framework is a free and open source Java library for developing and experimenting with multiobjective evolutionary algorithms (MOEAs) and other general-purpose optimization algorithms.  A number of algorithms are provided out-of-the-box, including NSGA-II, $\epsilon$-MOEA, GDE3 and MOEA/D.  In addition, the MOEA Framework provides the tools necessary to rapidly design, develop, execute and statistically test optimization algorithms.

This user manual is divided into the following three parts:
\begin{description}
  \item [Beginner's Guide] - Provides an introduction to the MOEA Framework for new users.  Topics discussed include installation instructions, walking through some introductory examples, and solving user-specified problems.
  \item [Advanced Guide] - Introduces features provided by the MOEA Framework intended for academic researchers and other advanced users.  Topics include performing large-scale experimentation, statistically comparing algorithms, and advanced configuration options.
  \item [Developer's Guide] - Intended for software developers, this part details guidelines for contributing to the MOEA Framework.  Topics covered include the development of new optimization algorithms, software coding guidelines and other policies for contributors.
\end{description}

\section{Key Features}

\begin{itemize}
  \item Fast, reliable implementations of many state-of-the-art multiobjective evolutionary algorithms
  \item Extensible with custom algorithms, problems and operators
  \item Modular design for constructing new optimization algorithms from existing components
  \item Permissive open source license
  \item Fully documented source code
  \item Over 1000 test cases to ensure validity
\end{itemize}

