% Copyright 2011-2014 David Hadka.  All Rights Reserved.
%
% This file is part of the MOEA Framework User Manual.
%
% Permission is granted to copy, distribute and/or modify this document under
% the terms of the GNU Free Documentation License, Version 1.3 or any later
% version published by the Free Software Foundation; with the Invariant Section
% being the section entitled "Preface", no Front-Cover Texts, and no Back-Cover
% Texts.  A copy of the license is included in the section entitled "GNU Free
% Documentation License".

\chapter{References}
\label{chpt:references}

Bleuler, S. et al.  ``PISA - A Platform and Programming Language Independent Interface for Search Algorithms.''  Evolutionary Multi-Criterion Optimization, pp. 494-508, 2003.
\newline
\newline
\noindent
Coello Coello, C.A., et al.  ``Evolutionary Algorithms for Solving Multi-Objective Problems.'' 2nd ed. New York, NY: Springer; 2007.
\newline
\newline
\noindent
Corne, D.W. and Knowles, J.D.  ``The Pareto-envelope based Selection Algorithm for Multiobjective Optimization.''  In Parallel Problem Solving from Nature (PPSN VI), pp. 839–848, 2000.
\newline
\newline
\noindent
Deb K., et al. ``Scalable Multi-objective Optimization Test Problems.'' In Proceedings of the Congress on Evolutionary Computation (CEC-2002); pp. 825–830, 2002.
\newline
\newline
\noindent
Deb, K. et al. ``A Fast Multi-Objective Evolutionary Algorithm for Finding Well-Spread Pareto-Optimal Solutions.'' KanGAL Report No 2003002, Feb 2003. 
\newline
\newline
\noindent
Deb, K. et al.  ``A Fast Elitist Multi-Objective Genetic Algorithm: NSGA-II.''  IEEE Transactions on Evolutionary Computation, 6:182-197, 2000.
\newline
\newline
\noindent
Deb, K. and Jain, H.  ``An Evolutionary Many-Objective Optimization Algorithm Using Reference-Point-Based Nondominated Sorting Approach, Part I: Solving Problems With Box Constraints.''  IEEE Transactions on Evolutionary Computation, 18(4):577-601, 2014.
\newline
\newline
\noindent
Durillo, J.J. and Nebro, A.J.  ``jMetal: a Java Framework for Multi-Objective Optimization.''  Advances in Engineering Software, 42:760-771, 2011.
\newline
\newline
\noindent
Durillo, J.J. et al.  ``Solving Three-Objective Optimization Problems Using a New Hybrid Cellular Genetic Algorithm.''  Parallel Problem Solving from Nature, pp. 661-670, 2008.
\newline
\newline
\noindent
Eskandari, H. et al.  ``FastPGA: A Dynamic Population Sizing Approach for Solving Expensive Multiobjective Optimization Problems.''  4th International Conference on Evolutionary Multi-Criterion Optimization, pp. 141-155, 2007.
\newline
\newline
\noindent
Goldberg, D.E.  ``Genetic Algorithms in Search, Optimization and Machine Learning.'' Reading, MA: Addison-Wesley Publishing Company; 1989.
\newline
\newline
\noindent
Hadka, D. and Reed, P. ``Borg: An Auto-Adaptive Many-Objective Evolutionary Computing Framework.'' Evolutionary Computation, 21(2):231-259, 2013.
\newline
\newline
\noindent
Hadka, D. and Reed, P. ``Diagnostic Assessment of Search Controls and Failure Modes in Many-Objective Evolutionary Optimization.'' Evolutionary Computation, 20(3):423-452, 2012.
\newline
\newline
\noindent
Horn, J. and  Nafpliotis, F.  ``Multiobjective Optimization using the Niched Pareto Genetic Algorithm.''  Tech. Rep. IlliGAL Report No. 93005, University of Illinois; 1993.
\newline
\newline
\noindent
Huband, S., et al.  ``A Review of Multiobjective Test Problems and a Scalable Test Problem Toolkit.''  IEEE Transactions on Evolutionary Computation, 10(5):477–506, 2006.
\newline
\newline
\noindent
Ishibuchi, H., et al.  ``Indicator-based Evolutionary Algorithm with Hypervolume Approximation by Achievement Scalarizing Functions.''  In Proceedings of the 12th Annual Conference on Genetic and Evolutionary Computation, GECCO ’10, pp. 527–534, 2010.
\newline
\newline
\noindent
Kollat, J. B., and Reed, P. M. ``Comparison of Multi-Objective Evolutionary Algorithms for Long-Term Monitoring Design.''  Advances in Water Resources, 29(6):792-807, 2006.
\newline
\newline
\noindent
Kukkonen and Lampinen. ``GDE3: The Third Evolution Step of Generalized Differential Evolution.'' KanGAL Report Number 2005013, 2005. 
\newline
\newline
\noindent
Li, H. and Zhang, Q.  ``Multiobjective Optimization problems with Complicated Pareto Sets, MOEA/D and NSGA-II.''  IEEE Transactions on Evolutionary Computation, 13(2):284-302, 2009.
\newline
\newline
\noindent
Nebro, A.J. et al.  ``AbYSS: Adapting Scatter Search to Multiobjective Optimization.''  IEEE Transactions on Evolutionary Computation.''  12(4), 2008.
\newline
\newline
\noindent
Nebro, A.J. et al.  ``Optimal Antenna Placement using a New Multi-objective CHC Algorithm.''  Proceedings of the 9th Annual Conference on Genetic and Evolutionary Computation, pp. 876-883, 2007.
\newline
\newline
\noindent
Nebro, A.J. et al.  ``A Cellular Genetic Algorithm for Multiobjective Optimization.''  Proceedings of the Workshop on Nature Inspired Cooperative Strategies for Optimization, pp. 25-36, 2006.
\newline
\newline
\noindent
Nebro, A.J. et al.  ``SMPSO: A New PSO-based Metaheuristic for Multi-objective Optimization.''  2009 IEEE Symposium on Computational Intelligence in Multicriteria Decision-Making, pp. 66-73, 2009.
\newline
\newline
\noindent
Reed, P., et al.  ``Evolutionary Multiobjective Optimization in Water Resources: The Past, Present, and Future.'' (Editor Invited Submission to the 35th Anniversary Special Issue), Advances in Water Resources, 51:438-456, 2013.
\newline
\newline
\noindent
Reyes, M. and Coello Coello, C.A.  ``Improving PSO-based Multi-objective Optimization using Crowding, Mutation and $\epsilon$-Dominance.''  Third International Conference on Evolutionary Multi-Criterion Optimization, pp. 509-519, 2005.
\newline
\newline
\noindent
Sierra, M.R. and Coello Coello, C.A.  ``Improving PSO-based Multi-objective Optimization using Crowding, Mutation and e-Dominance.''  Berlin: Springer; pp. 505–519, 2005.
\newline
\newline
\noindent
Saltelli, A., et al. ``Global Sensitivity Analysis: The Primer.'' Wiley, 2008.
\newline
\newline
\noindent
While, L. et al.  ``A Fast Way of Calculating Exact Hypervolumes.''  IEEE Transactions on Evolutionary Computation, 16(1):86-95, 2012.
\newline
\newline
\noindent
Zitzler, E., et al.  ``SPEA2: Improving the Strength Pareto Evolutionary Algorithm.''  In Evolutionary Methods for Design, Optimisation, and Control, Barcelona, Spain; pp. 95–100, 2002.
\newline
\newline
\noindent
Zitzler E. and K\"unzli, S.  ``Indicator-based Selection in Multiobjective Search.''  In Proceedings of the 8th International Conference on Parallel Problem Solving from Nature, pp. 832-842, 2004.
\newline
\newline
\noindent
Zhang, Q., et al. ``The Performance of a New Version of MOEA/D on CEC09 Unconstrained MOP Test Instances.'' IEEE Congress on Evolutionary Computation, pp. 203-208, 2009. 